---
title: "Reading and writing Quake WAL textures with wal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and writing Quake WAL textures with wal}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this document, we show how to read and write WAL textures, which are used by idtech1 and idtech2 games.


## Reading WAL files

If you only need the pixel data of a WAL file, try this:

```{r}
library("wal");
wal_file = system.file("extdata", "bricks.wal", package = "wal", mustWork = TRUE);
wal_image = wal::readWAL(wal_file);
```

The return value is an array with 3 dimensions, representing image width, height, and channels. It contains RGB color values in range 0..1:

```{r}
dim(wal_image);
```


To read the WAL file and get more detailed data, including the header and all mipmaps, read it into a wal instance instead:

```{r}
wal = wal::read.wal(wal_file);
```

This allows you to do more things, like converting to other formats and re-writing to other files. The wal instance is a named list, feel free to explore it. E.g., to see the header information, do this:

```{r}
wal$header$width;
```

### Preview a WAL texture in R

If you loaded a wal instance, you can plot it:

```{r}
plot(wal);
```

And with more control:

```{r}
plotwal.mipmap(wal, apply_palette = wal::pal_q2(), mip_level = 1)
```

The mipmaps are 0..4, where 0 is the largest (highest quality) version.

### Writing WAL images
